---
title: 'Week 7: Dates and Times'
author: "Olivia Guswiler"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Working with Dates and Strings

This week, we are going to learn a bit about how to work with some of the more troublesome data types in R: dates and character strings.

I've divided the lessons and the assignment into 2 parts this week. Today, we will work with the `lubridate` package to learn about working with dates and times. Part 1 of the Assignment will do the same.

## Set-Up

As is becoming standard for us now, let's start by loading in the `tidyverse`.

```{r}
library(tidyverse)
library(here)

# we will be using lubridate for part 1 and stringr for part 2
# for working without dates, the package hms is useful, but we won't be covering it in this class
```

We are going to use some stream data from the Santa Cruz River here in Tucson. The data are recorded every 15 minutes; these data are from 2023.

```{r}
download.file("https://nwis.waterservices.usgs.gov/nwis/iv/?sites=09482495&parameterCd=00060&startDT=2023-01-01T00:00:00.000-07:00&endDT=2023-12-31T23:59:59.999-07:00&siteStatus=all&format=rdb", here("data_raw/santa_cruz.txt"))
```

When we take a look at the data, we can see a number of rows that we need to skip when we read in the data...27 to be exact.

```{r}
water <- read_tsv(here("data_raw/santa_cruz.txt"))   # read in .txt files

# there are 27 rows of meta data, so we will specify in fn to exclude those rows

water <- read_tsv(here("data_raw/santa_cruz.txt"),
                  skip = 27, col_names = TRUE)
```

Ok, we are close but not quite there. The first row of data is metadata rather than data.

To remove that, we are going to use a new function called `slice()`, from the `dplyr` package. It allows us to select (or remove, in our case) rows from a dataframe based on position. We want to remove the first row.

```{r}
# we have our data and column names, but we have a first row full of measurement scale and units

water <- water %>% 
  slice(-1)         # use slice() to pull out specific rows in the data "1" means remove
head(water)
tail(water)
```

That's better! **Let's also rename the last 2 columns**. Those numbers are going to be pretty annoying to deal with down the road.

The stream discharge column should be numeric but got read in as character for some reason, so we need to convert it to a numeric data class.

```{r}
# rename last two columns

water <- water %>% 
  rename(discharge_ft3_s = "307121_00060",             # new name = old name
         qa_code = "307121_00060_cd") %>%              # use '' or "" around these, b/c the column name starts with a number
  mutate(discharge_ft3_s = as.double(discharge_ft3_s)) # discharge is currently in chr class, need it as numeric to do calculations

water
```

Let's try to plot the amount of water flowing through the stream through time.

```{r}
ggplot(water, aes(datetime, discharge_ft3_s)) +
  geom_point()
```

Well, that doesn't look great...

As it turns out, the `datetime` column is also currently character data. In order to use this column efficiently, we will want to convert it to a date (or datetime) data class.

## Dates and Times with `lubridate`

Dates and times can be particularly challenging to work with, but thankfully the `lubridate` package makes it a bit easier.

### Referencing the Current Date and Time

First, let's start with functions that return the current date and/or time. These are helpful if you want to "stamp" with the date/time something happens (e.g., when exactly did you knit that PDF file).

```{r}
# reference current date and time, useful to stamp with date and time it was produced

today()   # empty will give us date only

now()     # empty will give date and timezone in UTC

now(tzone = "MST") # date and time in MST
```

Aside from these functions, we will mostly be using the `lubridate` functions within a `mutate` function to modify entire columns at once.

### Making a `datetime` Column

The most generic function in `lubridate` is `as_date` or the variant, `as_datetime`. These convert non-dates into a date data class.

We need to specify the following arguments:

-   the data to be converted to a datetime
-   the format of the datetime data
-   optionally, we can specify the time zone

The way we specify the format of the datetime is a bit unusual. We use "conversion specifications," which are introduced by a `%` and are typically represented by a letter. There are lots of options, but the ones we will use are below:

-   `%Y`: 4 digit year
-   `%y`: 2 digit year
-   `%m`: month in numeric form
-   `%d`: day of month in numeric form
-   `%H`: hours in 24-hour format
-   `%M`: minutes
-   `%S`: seconds

The characters that are used to separate the date and time data are interpreted literally. What does that mean? If the date is separated by `-`, the conversion specifications should be separated by `-`, too.

For example, if our dates were formatted as `02/29/24`, the corresponding conversion format would be `%m/%d/%y`.

Let's convert our datetime column to a datetime data class.

```{r}
water %>% 
  mutate(datetime = as_datetime(datetime, format = "%Y-%m-%d %H:%M", tz = "MST"))

# the weird data type is stored in a way that allows us to perform calculations on dates and time in ways which would be very difficult otherwise

```

We can now see that the `datetime` column is something new, called `POSIXct`.

You don't need to know too much about what that means beyond that this is a common way datetimes are stored in R. Behind the scenes, the date/time is being stored as the number of seconds since Jan 1, 1970 at 00:00:00 UTC, allowing R to perform mathematical calculations with dates and times, which I'll demonstrate latter in the lesson.

### Functions Specifying Date/Time Structure

**Any time I try to remember the conversion specifications, I draw a blank. Are they uppercase or lowercase? Some letters don't match the first letter of the data it represents.**

`lubridate` provides functions with nearly every iteration and combination of year, month, day, hour, minutes, and seconds.

When we use these, we don't need to specify the exact format or the separators, which is particularly helpful if you have dates with different separators in the same column.

As we know from the previous code, our `datetime` column has year, month, day, hours and minutes.

```{r}
water <- water %>% 
  mutate(datetime = ymd_hm(datetime))  # does the samething as datetime, but a bit easier than remembering all those codes
water
```

Now, when we plot the trend of water discharge through 2023, we can see that ggplot knows how to deal with the values more efficiently.

```{r}
ggplot(water, aes(datetime, discharge_ft3_s)) +
  geom_point()

# it will now condense the dates in a readable format
```

#### Let's Practice

Work on Questions 1, 2, 4a and 4b

### Extracting Specific Components

Sometimes, we don't want all of the data from the datetime or we want to `filter`, `group_by`, `summarize`, etc. by a part of the datetime.

For example, perhaps we want to know the average water discharge per month. We would first want to create a new column with the month for each observation and then `group_by` and `summarize`.

There are numerous functions we can use to extract specific components from the datetime.

We can use the `date()` function to pull out the date from the datetime. As a note, it is surprisingly challenging to create a time column...

```{r}
water %>% 
  mutate(date = date(datetime))
```

Let's create columns for each segment.

```{r}
water <- water %>% 
  mutate(year = year(datetime),
         month = month(datetime),
         day = day(datetime),
         hour = hour(datetime),
         minute = minute(datetime),
         second = second(datetime))
water
```

Now, we can calculate the average value per month.

```{r}
water %>% 
  group_by(month) %>% 
  summarize(mean = mean(discharge_ft3_s))  # get calculations per month
```

We can also create new columns such as day of year (DOY). This is the day of year, running from 0-365. One common use for DOY is in phenology, measuring the annual timing of life history stages. The function to calculate DOY is `yday()`.

```{r}
tail(water) %>% 
  mutate(DOY = yday(datetime))
```

#### Let's Practice

Work on 4d (skip 4c for now)

### Creating Dates from Segments

What if you are given a dataset that has separate year, month, and day columns that you need to combine? If you recall, this is how the data frame for the Portal rodents is set up.

```{r}
surveys <- read_csv(here("data_raw/surveys.csv"))
surveys
```

We can bring these columns together to form a date column using the `make_date()` function.

```{r}
# skip using unite and changing data to date format by using make_date()

surveys <- surveys %>% 
  mutate(date = make_date(year = year,
                          month = month,
                          day = day),
         .before = "month")
surveys
```

### Performing Calculations

Because of the unique way that dates and datetimes are stored in R, we can also perform calculations with them.

In particular, `lubridate` has functions to calculate intervals between dates and datetimes.

```{r}
start <- min(surveys$date, na.rm = TRUE)
start

end <- max(surveys$date, na.rm = TRUE)
end

date_interval <- interval(start, end)
date_interval

as.duration(date_interval)
```

This lesson has really only scraped the surface of working with dates, but it should get you started if you need to do so in the future!

You should now be able to complete the remaining questions in the assignment.
