---
title: "Guswiler_Final"
author: "Olivia Guswiler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Temporal Ordering Memory Task. Data from Evelyn F. McKnight Brain Institute, University of Arizona. All rights and use of this data is reserved to the lab of Dr. Carol Barnes and is not available for reproduction or personal use.

# Setup

Install and load necessary packages.

```{r, echo = FALSE}
# install.packages("tidyverse")
library(tidyverse)
```

# Data Import and Make Tidy

## Import Data

Read .csv files into global environment and merge into a single data frame.

Include metadata file for information on each column.

```{r, echo = FALSE}
# Read in files
animal_info <- read_csv("../data_raw/animal_info.csv")
sample_1 <- read_csv("../data_raw/sample_1_obj_4.csv")
sample_2 <- read_csv("../data_raw/sample_2_obj_19.csv")
test <- read_csv("../data_raw/test.csv")

# Merge
full_data <- animal_info %>% 
  left_join(sample_1) %>% 
  left_join(sample_2) %>% 
  left_join(test) %>% 
  print()
```

## How to Proceed

There are several factors that need to be accounted for prior to beginning analyses:

-   Calculate Total Exploration during the first 2 minutes of the test phase and entire 4 minutes of the test phase.
    -   The first 2 minutes of exploration activity has been shown to be more indicative of recognition memory than exploration exhibited during later in the test phase, therefore we use only the first 2 minutes to infer recognition.
-   Determine each rat's Age Category based on their Age at Test.
    -   Young = 6 to 16.6 months
    -   Old = 16.7 months and up
-   Calculate each rat's Discrimination Ratio for the first 2 minutes of the test phase.
    -   Discrimination Ratio (DR): A metric commonly used to represent recognition memory sensitivity, based off on preferential exploration of an object that is less familiar over an object that is more familiar. A positive DR indicates preferential exploration for the less familiar object, a negative DR indicates preferential exploration of the familiar object, DR = 0 indicates no preference.
    -   DR = (t~object from sample 1~ -- t~object from sample 2~) / T~total~
        -   t = exploration time (s)
-   Rats need to meet an exploration criteria at each phase of the task to encode an event to their memory and recall it. Therefore, individuals with insufficient exploration could confound results if they didn't explore an object during a sample phase and experience that object as novel during the test phase and should be excluded from analyses. Minimum exploration criteria is as follows:
    -   Sample Phases: exploration time for [each]{.underline} object [\>]{.underline} 5 s
    -   Test Phase: total exploration time [\>]{.underline} 10 s

Here is the order of operations in which I will proceed to accomplish everything stated above:

1.  Add Age Category column
2.  Calculate total exploration during first 2 min of and entire test phase
3.  Calculate the discrimination ratios from exploration during first 2 min of test phase
4.  Create and apply a function that will indicate whether the rat had in/sufficient exploration during any phase of the task.

## Make Tidy

```{r}
full_data <- full_data %>% 
  # add age category
  mutate(age_cat =
           case_when(
             age_at_test_mo >= 16.7 ~ "old",
             age_at_test_mo < 16.7 ~ "young"),
         .before = "strain") %>% 
  # calculate test total exploration
  mutate(test_total_expl_s = test_expl_obj_4_s + test_expl_obj_19_s,
         .after = test_expl_obj_19_s) %>% 
  # calculate test first 2 min total exploration
  mutate(test_first_2min_total_expl_s =
           test_first_2min_expl_obj_4_s + test_first_2min_expl_obj_19_s,
         .after = test_first_2min_expl_obj_19_s) %>% 
  # calculate first 2 min DR
  mutate(disc_rat_2min = 
           (test_first_2min_expl_obj_4_s - test_first_2min_expl_obj_19_s) / test_first_2min_total_expl_s,
         .after = test_first_2min_total_expl_s) %>% 
  print()
```

Below, I've created two functions that will mark a rat either `TRUE` or `FALSE` for whether they exhibited sufficient exploration in sample phases 1 and 2 and the test phase.

```{r}
# Function to check for sufficient exploration from sample phases
check_sample_phase_expl <- function(expl_e,  # sample phase 1|2: time exploring East object
                                    expl_w){ # sample phase 1|2: time exploring West object
  sufficient_expl <- case_when(expl_e < 5 |
                                 expl_w < 5
                               ~ FALSE,
                               expl_e >= 5 |
                                 expl_w >= 5
                               ~ TRUE)
  return(sufficient_expl)
}


# Function to check for sufficient exploration from test phase
check_test_phase_expl <- function(test_total_expl){ # test: total time exploring both objects
  sufficient_expl <- case_when(test_total_expl < 10
                               ~ FALSE,
                               test_total_expl >= 10
                               ~ TRUE)
  return(sufficient_expl)
}
```

After loading the functions into the environment, I use them to create new columns for each phase of the task and then merge these columns into a single column to indicate sufficient exploration. Because a single instance of insufficient exploration at any phase results in the exclusion of a rat, the merged column will retain `FALSE` unless all indicate `TRUE`.

```{r}
full_data <- full_data %>%
  mutate(.after = notes,
         # check sample phase 1 exploration
         sample_1_suff_expl = check_sample_phase_expl(sample_1_east_obj_4_s,
                                                      sample_1_west_obj_4_s),
         # check sample phase 2 exploration
         sample_2_suff_expl = check_sample_phase_expl(sample_2_east_obj_19_s,
                                                      sample_2_west_obj_19_s),
         # check test exploration
         test_suff_expl = check_test_phase_expl(test_total_expl_s)) %>% 
  # merge all T/F columns into one, marking F unless all are T
  mutate(.before = sample_1_suff_expl,
         sufficient_expl = case_when(sample_1_suff_expl == TRUE &
                                       sample_2_suff_expl == TRUE &
                                       test_suff_expl == TRUE
                                     ~ TRUE,
                                     .default = FALSE)) %>% 
  # remove individual T/F columns used to create merged column
  select(!c(sample_1_suff_expl, sample_2_suff_expl, test_suff_expl)) %>% 
  print()
```

Save this data to the data_clean/ folder for ease of reference.

```{r}
write_csv(full_data, "../data_clean/tor_full_data.csv")
```

# Looking at the Data

To prevent the workspace from getting cluttered, I clear the environment below before reading in the data just saved above. While loading the data frame, remove rats with insufficient data, select columns relevant to analyses, and create groups for comparisons.

```{r}
rm(list = ls())

# read in .csv
analysis_data <- read_csv("../data_clean/tor_full_data.csv") %>% 
  # remove rats with insufficient exploration
  filter(sufficient_expl == TRUE) %>% 
  # select columns relevant to analysis
  select(barnes_id, age_cat, sex,
         disc_rat_2min,
         test_first_2min_total_expl_s, test_total_expl_s,
         test_first_visit_obj_4,
         sample_1_east_obj_4_s, sample_1_west_obj_4_s,
         sample_2_east_obj_19_s, sample_2_west_obj_19_s) %>% 
  # group by age category and sex
  group_by(age_cat, sex)
```

## Summarize

Before blindly running tests to make comparisons among groups, look at some summaries of the raw data to provide a guide.

```{r}
# Quick look at group summary data
analysis_data %>%
  summarise(n(),
            # DR
            mean(disc_rat_2min),
            sd(disc_rat_2min),
            # % of group that first approached the object from sample 1
            (sum(test_first_visit_obj_4) / n()) * 100,
            # total time exploring during test
            mean(test_total_expl_s),
            median(test_total_expl_s),
            sd(test_total_expl_s))
```

## Visualize

Right away, it can be seen that there appears to be a greater difference in performance between age rather than sex. Visualize the data before going forward.

```{r}
# Discrimination ratio box plot
analysis_data %>%
  ggplot(aes(age_cat, disc_rat_2min,
             # fill colors by sex
             fill = sex)) +
  geom_boxplot() +
  # add points by age and sex over corresponding box
  geom_point(position = position_dodge(width = 0.75)) +
  # add line at y = 0
  geom_abline(slope = 0,
              intercept = 0.0,
              linetype = "dotted") +
  # change y-axis limit to -1.0:1.0
  expand_limits(y = 1.0) +
  # change main and axes titles
  labs(title = "TOR Test Performance",
       x = "Age Category",
       y = "Discrimination Ratio") +
  # change legend title
  guides(fill = guide_legend(title = "Sex")) +
  theme_classic()

# Total exploration bar plot
analysis_data %>% 
  ggplot(aes(age_cat, test_total_expl_s,
             fill = sex)) +
  geom_bar(position = "dodge",
           stat = "identity",
           width = 0.75) +
  # change main and axes titles
  labs(title = "TOR Total Test Exploration",
       x = "Age Category",
       y = "Exploration (s)") +
  # change legend title
  guides(fill = guide_legend(title = "Sex")) +
  theme_classic()
```

There appears to be similar test performance and exploration between sexes, within age categories. Perform some analyses to confirm this.

# Analyses

To make the process of comparisons among groups easier, first split up the data into data frames based on age and sex.

```{r}
# Obtain the group names that will become the data frame names
naming <- unite(group_keys(analysis_data),
                "group", age_cat, sex)

# Create the data frames
by_group_data <- analysis_data %>% 
  # split the original data frames by groupings
  group_split() %>% 
  # name them based on the vector we created earlier
  set_names(nm = naming$group) %>% 
  # add these to our working environment
  list2env(envir = .GlobalEnv)

head(old_F)
head(old_M)
head(young_F)
head(young_M)

# Save these split groups in case of future needs
#### keep this??
write_csv(old_F, "../data_clean/tor_old_female.csv")
write_csv(old_M, "../data_clean/tor_old_male.csv")
write_csv(young_F, "../data_clean/tor_young_female.csv")
write_csv(young_M, "../data_clean/tor_young_male.csv")
```

Perform t-tests to compare differences between sexes, within the two age categories.

```{r}
# Young rats
#  t-test comparing performance during test phase
t.test(young_F$disc_rat_2min,
       young_M$disc_rat_2min,
       # variance is relatively the same between groups
       var.equal = TRUE)

#  t-test comparing total exploration during test phase
t.test(young_F$test_total_expl_s,
       young_M$test_total_expl_s,
       var.equal = TRUE)

# Old rats
#  t-test comparing performance during test phase
t.test(old_F$disc_rat_2min,
       old_M$disc_rat_2min,
       var.equal = TRUE)

#  t-test comparing total exploration during test phase
t.test(old_F$test_total_expl_s,
       old_M$test_total_expl_s,
       var.equal = TRUE)
```

The results indicate that the null hypothesis cannot be rejected as false, within age categories the rats exhibit similar performance and exploration during the test phase, regardless of sex.

Now, make comparisons between age categories alone. Begin by clearing environment, reading in data again, then splitting into data frames based on age category.

```{r}
#b/c there is no difference between sexes, we can collapse these groups and make comparisons between age_cat
young <- bind_rows(young_F, young_M)
old <- bind_rows(old_F, old_M)


# remove below this in code chunk







rm(list = ls())

# read in .csv
analysis_data <- read_csv("../data_clean/tor_full_data.csv") %>% 
  # remove rats with insufficient exploration
  filter(sufficient_expl == TRUE) %>% 
  # select columns relevant to analysis
  select(barnes_id, age_cat, sex,
         disc_rat_2min,
         test_first_2min_total_expl_s, test_total_expl_s,
         test_first_visit_obj_4,
         sample_1_east_obj_4_s, sample_1_west_obj_4_s,
         sample_2_east_obj_19_s, sample_2_west_obj_19_s) %>% 
  # group by age category
  group_by(age_cat)

# Obtain the group names that will become the data frame names
naming <- unite(group_keys(analysis_data),
                "group", age_cat)

# Create the 'young' and 'old' data frames
by_group_data <- analysis_data %>% 
  # split the original data frames by groupings
  group_split() %>% 
  # name them based on the vector we created earlier
  set_names(nm = naming$group) %>% 
  # add these to our working environment
  list2env(envir = .GlobalEnv)

head(old)
tail(old)
head(young)
tail(young)
```

Visualize again

```{r}
# Discrimination ratio box plot
analysis_data %>%
  ggplot(aes(age_cat, disc_rat_2min,
             # fill colors by age category
             fill = age_cat)) +
  geom_boxplot() +
  # add points by age and sex over corresponding box
  geom_point(position = position_dodge(width = 0.75)) +
  # add line at y = 0
  geom_abline(slope = 0,
              intercept = 0.0,
              linetype = "dotted") +
  # change y-axis limit to -1.0:1.0
  expand_limits(y = 1.0) +
  # change main and axes titles
  labs(title = "TOR Test Performance",
       x = "Age Category",
       y = "Discrimination Ratio") +
  # set colors so they are different from previous plots
  scale_fill_brewer(palette = "Set3") +
  # remove legend
  theme(legend.position = "none") +
  theme_classic()

# Total exploration bar plot
analysis_data %>% 
  ggplot(aes(age_cat, test_total_expl_s,
             fill = age_cat)) +
  geom_bar(position = "dodge",
           stat = "identity",
           width = 0.75,
           # remove legend (needs to be in a different spot for some reason)
           show.legend = FALSE) +
  # change main and axes titles
  labs(title = "TOR Total Test Exploration",
       x = "Age Category",
       y = "Exploration (s)") +
  scale_fill_brewer(palette = "Set3") +
  theme_classic()
```

Now run analyses for comparison between age category.

```{r}

```
