---
title: "Guswiler_Final"
author: "Olivia Guswiler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Temporal Ordering Memory task. Data from Evelyn F. McKnight Brain Institute, University of Arizona. All rights and use of this data is reserved to the lab of Dr. Carol Barnes and is not available for reproduction or personal use.

## Setup

Install and load necessary packages.

```{r}
# Install required packages if you do not already have them
# install.packages("tidyverse")

library(tidyverse)
```

## Data Import

```{r, echo = FALSE}
# Read .csv files into workspace environment

```

Merge the data into one data frame.

```{r}

```

Observed issues:

-   Repetition of strings in the same cell, most often `General:General`.

-   Behavior categorized as `General` with the inclusion of another behavior.

-   One instance of non-conforming value `Sleep` == `Sleeping`

Decided order of operation:

1.  

```{r}

```

Remove columns containing unnecessary or repeated information.

```{r}


# Save this cleaned data for easy reference

```

If you wish to clean up your environment prior to continuing you can clear the console and environment below and read in the cleaned data file.

## Looking at the Data
